---
title: "Halloween Project"
author: "A17576411"
format: pdf
---
```{r}
candy_file <- "candy-data.csv"

candy = read.csv("candy-data.csv", row.names=1)
head(candy)
```
>Q1: How many different candy types are in this dataset?

A: 85 types of candy
```{r}
dim(candy)
```
>Q2: How many fruity candy types are in the dataset?

A: 38 fruity candy type
```{r}
sum(candy$fruity)
```
```{r}
candy["Twix", ]$winpercent
```
>Q3: What is your favorite candy in the dataset and what's its winpercent value?

A: 46.78%
```{r}
candy["Gobstopper", ]$winpercent
```

>Q4: What is the winpercent value for "Kit Kat"?

A: 76.77%
```{r}
candy["Kit Kat", ]$winpercent
```



>Q5: What is the winpercent value for "Tootsie Roll Snack Bars"?

A:49.65%
```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```
```{r}
#install.packages("skimr")
```

```{r}
#install.packages("skimr")
library("skimr")
skim(candy)
```

>Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

A: Winpercent

>Q7: What do you think a zero and one represent for the candy$chocolate column?

```{r}
skim(candy$chocolate)
```
A zero means the candy does not have chocolate

A one means the candy has chocolate

>Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
hist(candy$winpercent)

```

>Q9. Is the distribution of winpercent values symmetrical?

A: Judging from the graph, no.

>Q10. Is the center of the distribution above or below 50%?

A: above 50%

>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

A: It is

```{r}
mean(candy$winpercent[as.logical(candy$chocolate)]) > mean(candy$winpercent[as.logical(candy$fruity)])
```

>Q12. Is this difference statistically significant?

A: P value is 2.871e-08 <<< 0.05, so the difference is statistically significant
```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)], candy$winpercent[as.logical(candy$fruity)])
```

>Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```


>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
library(dplyr)
candy %>%
  arrange(desc(candy$winpercent)) %>%
  head(5)
```

>Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(x= candy$winpercent, y= rownames(candy)) +
  geom_bar(stat="identity")
```
>Q16:  This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(x= candy$winpercent, reorder(rownames(candy), winpercent)) +
  geom_bar(stat="identity")
```
```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

>Q17. What is the worst ranked chocolate candy?

Sixlets from the graph

>Q18. What is the best ranked fruity candy?

Starburst

```{r}
#install.packages("ggrepel")
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

A: Reese's Miniatures

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

A: Nik L Nip is the least popular
```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

>Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```




```{r}
#install.packages("corrplot")
library(corrplot)
```
```{r}
cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

A: fruity and chocolate

>Q23. Similarly, what two variables are most positively correlated?

A: It appears to be chocolate and bar


```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```
>Q24: What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

A: fruity, hard, and pluribus are picked up strongly by PC1 in the positive direction. This makes sense, as popular candies like Starburst, Nerds, and Skittles, fulfill many of these qualities if not all three.
